Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-12-24 survey/survey.tex
more on data III

\documentclass{article}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{quotes}

@@

In the previous sections, we examined equivalences between
conventional data structures, i.e, sets of values and structured trees
of values. We now consider a richer, less conventional, but
foundational notion of data: programs themselves. Indeed, universal
computation models fundamentally rely on the fact that ``programs are
(or can be encoded as) data'', e.g., a Turing machine can be encoded
as a string that another Turing machine (or even the same machine) can
manipulate. In our setting, the programs developed in the previous
sections are reversible deformations of data structures. Can these
reversible deformations be themselves be subject to (higher-level)
reversible deformations?

Before developing the theory, let's consider a small example:

\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
  \draw[<->,double,red,thick] (2.25,-1.5) -- (2.25,-2.5) ;
  \node at (3.3,-2) {$\mathit{swapl₊⇔}$} ;
  \node at (2.5,-1.3) {$((c₂~⊕~c₁)~◎~\mathit{swap₊})$};
  \node at (2.5,-2.7) {$(\mathit{swap₊}~◎~(c₁~⊕~c₂))$};
  \draw (-2,-2) ellipse (0.5cm and 1cm);
  \draw[fill] (-2,-1.5) circle [radius=0.025];
  \node[below] at (-2.1,-1.5) {$A$};
  \draw[fill] (-2,-2.5) circle [radius=0.025];
  \node[below] at (-2.1,-2.5) {$B$};

@@

  \draw (-2,-1.5) to[bend left] (1,0.5) ;
  \draw (-2,-2.5) to[bend left] (1,-0.5) ;
  \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
  \draw[->] (3.5,-0.5) to[bend left] (6.5,-2.45) ;

  \draw (-2,-1.5) to[bend right] (1,-3.5) ;
  \draw (-2,-2.5) to[bend right] (1,-4.5) ;
  \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
  \draw[->] (3.5,-4.5) to[bend right] (6.5,-2.55) ;

@@

  \node at (3,-3.5) {c₁};
  \node at (3,-4.5) {c₂};

  \node at (1.5,0.5) {c₂};
  \node at (1.5,-0.5) {c₁};

@@

\end{tikzpicture}
\end{center}

@@

>>>>

\documentclass{article}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{quotes}
\tikzset{>=latex}

@@

In the previous sections, we examined equivalences between
conventional data structures, i.e, sets of values and structured trees
of values. We now consider a richer, less conventional, but
foundational notion of data: programs themselves. Indeed, universal
computation models fundamentally rely on the fact that \emph{programs
  are (or can be encoded as) data}, e.g., a Turing machine can be
encoded as a string that another Turing machine (or even the same
machine) can manipulate. In our setting, the programs developed in the
previous section are reversible deformations between structured finite
types. We now ask whether these reversible deformations can themselves
be subject to (higher-level) reversible deformations?

Before developing the theory, let's consider a small example
consisting of two deformations between the types $A + B$ and $C+D$:

\begin{center}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
%%  \node at (3.3,-2) {$\mathit{swapl}_{+\Leftrightarrow}$} ;
%%  \node at (2.5,-1.3) {$((c_2~\oplus~c_1)~\circledcirc~\mathit{swap}_{+})$}; 
%%  \node at (2.5,-2.7) {$(\mathit{swap}_{+}~\circledcirc~(c_1~\oplus~c_2))$}; 
  \draw (-2,-2) ellipse (0.5cm and 1cm);
  \draw[fill] (-2,-1.5) circle [radius=0.025];
  \node[below] at (-2.1,-1.5) {$A$};
  \draw[fill] (-2,-2.5) circle [radius=0.025];
  \node[below] at (-2.1,-2.5) {$B$};

@@

  \draw[<-] (-2,-1.5) to[bend left] (1,0.5) ;
  \draw[<-] (-2,-2.5) to[bend left] (1,-0.5) ;
  \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
  \draw[->] (3.5,-0.5) to[bend left] (6.5,-2.45) ;

  \draw[<-] (-2,-1.5) to[bend right] (1,-3.5) ;
  \draw[<-] (-2,-2.5) to[bend right] (1,-4.5) ;
  \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
  \draw[->] (3.5,-4.5) to[bend right] (6.5,-2.55) ;

@@

  \node at (3,-3.5) {$c_1$};
  \node at (3,-4.5) {$c_2$};

  \node at (1.5,0.5) {$c_2$};
  \node at (1.5,-0.5) {$c_1$};

@@

\end{tikzpicture}
\end{center}
The top path is the reversible deformation
$(c_2~\oplus~c_1)~\circledcirc~\mathit{swap}_{+}$ which deforms the
type $A$ by $c_2$, deforms the type $B$ by $c_1$, and deforms the
resulting space by a twist that exchanges the two injections into the
sum type. The bottom path performs the twist first and then deforms
the type $A$ by $c_2$ and the type $B$ by $c_1$ as before. If one
could imagine the paths are physical wires and the deformations $c_1$
and $c_2$ as arbitrary deformations on these wires then, holding the
points $A$, $B$, $C$, and $D$ fixed, it is possible to rotate the top
part of the diagram to become identical to the bottom one. In other
words, there exists a higher-level deformation of the program
$(c_2~\oplus~c_1)~\circledcirc~\mathit{swap}_{+}$ to the program
$\mathit{swap}_{+} \circledcirc (c_1~\oplus~c_2)$.

@@

\begin{figure}[t]
Let $c₁ : t₁ ⟷ t₂$,  $c₂ : t₂ ⟷ t₃$, and $c₃ : t₃ ⟷ t₄$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {c₁ ◎ (c₂ ◎ c₃) ⇔ (c₁ ◎ c₂) ◎ c₃}
\\
  {(c₁ ⊕ (c₂ ⊕ c₃)) ◎ \assoclp ⇔ \assoclp ◎ ((c₁ ⊕ c₂) ⊕ c₃)}
\\
  {(c₁ ⊗ (c₂ ⊗ c₃)) ◎ \assoclt ⇔ \assoclt ◎ ((c₁ ⊗ c₂) ⊗ c₃)}
\\
  {((c₁ ⊕ c₂) ⊕ c₃) ◎ \assocrp ⇔ \assocrp ◎ (c₁ ⊕ (c₂ ⊕ c₃))}
\\
  {((c₁ ⊗ c₂) ⊗ c₃) ◎ \assocrt ⇔ \assocrt ◎ (c₁ ⊗ (c₂ ⊗ c₃))}
\\
  {\assocrp ◎ \assocrp ⇔ ((\assocrp ⊕ \idc) ◎ \assocrp) ◎ (\idc ⊕ \assocrp)}
\\
  {\assocrt ◎ \assocrt ⇔ ((\assocrt ⊗ \idc) ◎ \assocrt) ◎ (\idc ⊗ \assocrt)}
\end{array}\]
\caption{\label{figj}Signatures of level-1 $\Pi$-combinators: associativity}
\end{figure}

\begin{figure}[t]
Let $a : t₁ ⟷ t₂$, $b : t₃ ⟷ t₄$, and $c : t₅ ⟷ t₆$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {((a ⊕ b) ⊗ c) ◎ \dist ⇔ \dist ◎ ((a ⊗ c) ⊕ (b ⊗ c))}
\\
  {(a ⊗ (b ⊕ c)) ◎ \distl ⇔ \distl ◎ ((a ⊗ b) ⊕ (a ⊗ c))}
\\
  {((a ⊗ c) ⊕ (b ⊗ c)) ◎ \factor ⇔ \factor ◎ ((a ⊕ b) ⊗ c)}
\\
  {((a ⊗ b) ⊕ (a ⊗ c)) ◎ \factorl ⇔ \factorl ◎ (a ⊗ (b ⊕ c))}
\end{array}\]
\caption{\label{figi}Signatures of level-1 $\Pi$-combinators: distributivity and factoring}
\end{figure}

\begin{figure}[t]
Let $c, c₁, c₂, c₃ : t₁ ⟷ t₂$ and $c', c'' : t₃ ⟷ t₄$: 
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\idc ◎ \, c ⇔ c}
\quad 
  {c \, ◎ \idc \, ⇔ c}
\quad
  {c\,\, ◎\,! c ⇔ \idc}
\quad 
  {! c ◎ c ⇔ \idc}
\\
  {c ⇔ c}
\quad 
\Rule{}
  {c₁ ⇔ c₂ \quad c₂ ⇔ c₃}
  {c₁ ⇔ c₃}
  {} 
\quad
\Rule{}
  {c₁ ⇔ c' \quad c₂ ⇔ c''}
  {c₁ ◎ c₂ ⇔ c' ◎ c''}
  {}
\end{array}\]
\caption{\label{figh}Signatures of level-1 $\Pi$-combinators: identity and composition}
\end{figure}

\begin{figure}[t]
Let $c₀ : 0 ⟷ 0$, $c₁ : 1 ⟷ 1$, and $c : t₁ ⟷ t₂$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\identlp ◎ c ⇔ (c₀ ⊕ c) ◎ \identlp}
\qquad 
  {\identrp ◎ (c₀ ⊕ c) ⇔ c ◎ \identrp}
\\
  {\identlsp ◎ c ⇔ (c ⊕ c₀) ◎ \identlsp}
\qquad
  {\identrsp ◎ (c ⊕ c₀) ⇔ c ◎ \identrsp}
\\
  {\identlt ◎ c ⇔ (c₁ ⊗ c) ◎ \identlt}
\qquad
  {\identrt ◎ (c₁ ⊗ c) ⇔ c ◎ \identrp}
\\
  {\identlst ◎ c ⇔ (c ⊗ c₁) ◎ \identlst}
\qquad
  {\identrst ◎ (c ⊗ c₁) ⇔ c ◎ \identrst}
\\
  {\identlt ⇔ \distl ◎ (\identlt ⊕ \identlt)}
\\
\identlp ⇔ \swapp ◎ \identrsp
\qquad
\identlt ⇔ \swapt ◎ \identrst
\end{array}\]
\caption{\label{figg}Signatures of level-1 $\Pi$-combinators: unit}
\end{figure}

\begin{figure}[t]
Let $c₁ : t₁ ⟷ t₂$ and $c₂ : t₃ ⟷ t₄$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\swapp ◎ (c₁ ⊕ c₂) ⇔ (c₂ ⊕ c₁) ◎ \swapp}
\quad
  {\swapt ◎ (c₁ ⊗ c₂) ⇔ (c₂ ⊗ c₁) ◎ \swapt}
\\
  {(\assocrp ◎ \swapp) ◎ \assocrp ⇔ ((\swapp ⊕ \idc) ◎ \assocrp) ◎ (\idc ⊕ \swapp)}
\\
  {(\assoclp ◎ \swapp) ◎ \assoclp ⇔ ((\idc ⊕ \swapp) ◎ \assoclp) ◎ (\swapp ⊕ \idc)}
\\
  {(\assocrt ◎ \swapt) ◎ \assocrt ⇔ ((\swapt ⊗ \idc) ◎ \assocrt) ◎ (\idc ⊗ \swapt)}
\\
  {(\assoclt ◎ \swapt) ◎ \assoclt ⇔ ((\idc ⊗ \swapt) ◎ \assoclt) ◎ (\swapt ⊗ \idc)}
\end{array}\]
\caption{\label{figf}Signatures of level-1 $\Pi$-combinators: commutativity and associativity}
\end{figure}

\begin{figure}[t]
Let $c₁ : t₁ ⟷ t₂$, $c₂ : t₃ ⟷ t₄$, $c₃ : t₁ ⟷ t₂$, and $c₄ : t₃ ⟷ t₄$. \\
Let $a₁ : t₅ ⟷ t₁$,  $a₂ : t₆ ⟷ t₂$, $a₃ : t₁ ⟷ t₃$, and $a₄ : t₂ ⟷ t₄$.
\[\def\arraystretch{1.3}
\begin{array}{c}
\Rule{}
  {c₁ ⇔ c₃ \quad c₂ ⇔ c₄}
  {c₁ ⊕ c₂ ⇔ c₃ ⊕ c₄}
  {}
\qquad
\Rule{}
  {c₁ ⇔ c₃ \quad c₂ ⇔ c₄}
  {c₁ ⊗ c₂ ⇔ c₃ ⊗ c₄}
  {} 
\\
  {\idc ⊕ \, \idc \, ⇔ \idc}
\qquad
  {\idc ⊗ \, \idc \, ⇔ \idc}
\\
  {(a₁ ◎ a₃) ⊕ (a₂ ◎ a₄) ⇔ (a₁ ⊕ a₂) ◎ (a₃ ⊕ a₄)}
\\
  {(a₁ ◎ a₃) ⊗ (a₂ ◎ a₄) ⇔ (a₁ ⊗ a₂) ◎ (a₃ ⊗ a₄)}
\end{array}\]
\caption{\label{fige}Signatures of level-1 $\Pi$-combinators: functors}
\end{figure}

\begin{figure}[t]
\[\def\arraystretch{1.3}
\begin{array}{c}
  {\identlsp ⊕ \idc ~⇔~ \assocrp ◎ (\idc ⊕ \, \identlp)}
\\
  {\identlst ⊗ \idc ~⇔~ \assocrt ◎ (\idc ⊗ \, \identlt)}
\end{array}\]
\caption{\label{figd}Signatures of level-1 $\Pi$-combinators: unit and associativity}
\end{figure}

\begin{figure}[t]
Let $c : t₁ ⟷ t₂$:
\[\def\arraystretch{1.3}
\begin{array}{c}
  {(c ⊗ \idc) ◎ \absorbl ⇔ \absorbl ◎ \idc}
\quad
  {(\idc \, ⊗ c) ◎ \absorbr ⇔ \absorbr ◎ \idc}
\\
  {\idc ◎ \, \factorzl ⇔ \factorzl ◎ (\idc ⊗ c)}
\quad
  {\idc ◎ \, \factorzr ⇔ \factorzr ◎ (c ⊗ \idc)}
\\
  {\absorbr ⇔ \absorbl}
\\
  {\absorbr ⇔ (\distl ◎ (\absorbr ⊕ \absorbr)) ◎ \identlp}
\\
  {\identlst ⇔ \absorbr}
\qquad
  {\absorbl ⇔ \swapt ◎ \absorbr}
\\
  {\absorbr ⇔ (\assoclt ◎ (\absorbr ⊗ \idc)) ◎ \absorbr}
\\
  {(\idc ⊗ \absorbr) ◎ \absorbl ⇔ (\assoclt ◎ (\absorbl ⊗ \idc)) ◎ \absorbr}
\\
  {\idc ⊗ \, \identlp ⇔ (\distl ◎ (\absorbl ⊕ \idc)) ◎ \identlp}
\end{array}\]
\caption{\label{figc}Signatures of level-1 $\Pi$-combinators: zero}
\end{figure}

\begin{figure}[t]
\[\def\arraystretch{1.3}
\begin{array}{c}
  {((\assoclp ⊗ \idc) ◎ \dist) ◎ (\dist ⊕ \idc) ⇔ (\dist ◎ (\idc ⊕ \dist)) ◎ \assoclp}
\\
  {\assoclt ◎ \distl ⇔ ((\idc ⊗ \distl) ◎ \distl) ◎ (\assoclt ⊕ \assoclt)}
\end{array}\]
\vspace{ -0.5em}
\[\def\arraystretch{1.3}
\begin{array}{rcl}
  (\distl ◎ (\dist ⊕ \dist)) ◎ \assoclp &⇔&   
   \dist ◎ (\distl ⊕ \distl) ◎ \assoclp ~◎ \\
&& (\assocrp ⊕ \idc) ~◎ \\
&& ((\idc ⊕ \swapp) ⊕ \idc) ~◎ \\
&&      (\assoclp ⊕ \idc)
\end{array}\]
\caption{\label{figb}Signatures of level-1 $\Pi$-combinators: associativity and distributivity}
\end{figure}

\begin{figure}[t]
\[\def\arraystretch{1.3}
\begin{array}{rcl}
  (\idc ⊗ \swapp) ◎ \distl &⇔& \distl ◎ \swapp
\\
  \dist ◎ (\swapt ⊕ \swapt) &⇔& \swapt ◎ \distl
\end{array}\]
\caption{\label{figa}Signatures of level-1 $\Pi$-combinators: commutativity and distributivity}
\end{figure}
